UKB Raw Analyses

### 2023-12-05 eda/ukb_raw_analyses.Rmd
split out raw ukb_CN analyses from analyses of combat effects

---
title: "UKB Simulation Analyses"
output: html_document
date: '2023-10-06'
---

@@

source("./cubic_scripts/R_scripts/gamlss_helper_funs.R")
# 
# # Enable this universe
# options(repos = c(
#   ggseg = 'https://ggseg.r-universe.dev',
#   CRAN = 'https://cloud.r-project.org'))
# 
# # Install some packages
# install.packages('ggsegExtra')
```

@@

```{r}
ukb_df_totals <- read_csv(file="data/ukb_CN_data.csv")
ukb_df_age.filt <- read_csv(file="data/ukb_to_model/ukb_CN_data_simsites.csv")

vol_list_global <- readRDS(file="cubic_scripts/R_scripts/vol_list_global.rds")
vol_list_regions <- readRDS(file="cubic_scripts/R_scripts/Vol_list_regions.rds")
sa_list <- readRDS(file="cubic_scripts/R_scripts/SA_list.rds")
ct_list <- readRDS(file="cubic_scripts/R_scripts/CT_list.rds")
pheno_list <- readRDS(file="cubic_scripts/R_scripts/pheno_list.rds")

aseg_cerebrum <- readRDS(file="/Users/megardn/Desktop/BGD_Repos/ggseg_atlas/aseg_cerebrum_atlas.RDS")
```

@@

```{r}
drop1.df <- fread("ukb_basic/drop1_tests.csv", stringsAsFactors = TRUE) %>%
  dplyr::select(-V1)
sex.df <- fread("ukb_basic/sex_summary.csv", stringsAsFactors = TRUE) %>%
  dplyr::select(-V1)
sum.df <- fread("ukb_basic/gamlss_summary.csv", stringsAsFactors = TRUE) %>%
  dplyr::select(-V1) %>%
  mutate(dataset = gsub("-data", "", dataset)) %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb-CN-simsites"))
```

@@

Simple bar plot of sites
```{r}
ggplot(ukb_df_age.filt) +
  geom_bar(aes(x=sim.site, fill=sex), color="black", alpha=0.6) +
  theme_linedraw(base_size=12)
```

## Raw Data Results
Review outcomes from raw (non-ComBat-ed) data

@@

## ComBat Effects by Moment

Loading combatted data (scp'd from cubic: `scp -r 'gardnerm@cubic-login.uphs.upenn.edu:/cbica/home/gardnerm/combat-biovar/data/ukb_to_model/*.csv' ./data/ukb_to_model`) to see how values change with different combat methods

```{r}
data.csvs <- list.files(path = "data/ukb_to_model", pattern = ".csv", full.names = TRUE)
combined_data <- data.frame()

for (file in data.csvs) {
  # Read each CSV file
  data <- fread(file)

  # Add a "Source_File" column with the file name
  data <- data %>%
    mutate(Source_File = as.factor(basename(file)))

  # Bind the data to the combined dataframe
  combined_data <- bind_rows(combined_data, data, .id = "File_ID")
}

table(combined_data$Source_File)
```

```{r}
var.df <- combined_data %>%
  group_by(Source_File, sim.site) %>%
  summarize(across(all_of(pheno_list), var))

var.long <- pivot_longer(var.df, cols=c(-Source_File, -sim.site), names_to = "feature", values_to = "variance")
```

```{r}
var.long %>%
  dplyr::filter(!(feature %in% vol_list_global)) %>%
  ggplot() +
  geom_density(aes(x=variance, fill=Source_File), alpha = 0.4, position="identity") +
  facet_wrap(~sim.site)

var.long %>%
  dplyr::filter(!(feature %in% vol_list_global) & (Source_File == "cf.gam_data.csv" |Source_File == "ukb_CN_data_simsites.csv" |  Source_File == "cf.gamlss_data.csv" )) %>%
  ggplot() +
  geom_density(aes(x=variance, fill=Source_File), alpha = 0.5, position="identity") +
  facet_wrap(~sim.site)

#Source_File == "cf.lm_data.csv" | 
```

```{r}
var.by.sex <- combined_data %>%
  group_by(Source_File, sex) %>%
  summarize(across(all_of(pheno_list), var))

var.by.sex.long <- pivot_longer(var.by.sex, cols=c(-Source_File, -sex), names_to = "feature", values_to = "variance")
```

```{r}
var.by.sex.long %>%
  dplyr::filter(!(feature %in% vol_list_global) & (Source_File == "cf.gam_data.csv" | Source_File == "cf.gamlss_data.csv" | Source_File == "ukb_CN_data_simsites.csv")) %>%
  ggplot() +
  geom_density(aes(x=variance, fill=sex), alpha = 0.5, position="identity") +
  facet_wrap(~Source_File)
```

Simple t test for sex effects across combat types
```{r}
sex_ttest_df <- data.frame("Source_File" = levels(combined_data$Source_File))

for(pheno in pheno_list){
  # New column name
  new_col_name <- paste0(pheno, ".p")

  #run t-tests
  out <- lapply(split(combined_data, factor(combined_data$Source_File)), function(x)t.test(x[[pheno]] ~ x[["sex"]], paired=FALSE)$p.value)
  stopifnot(names(out) == levels(combined_data$Source_File)) #make sure results are in correct order

  #append to dataframe
  sex_ttest_df[[new_col_name]] <- unlist(unname(out))
}
```

```{r}
sex_ttest_long <- sex_ttest_df %>% pivot_longer(!Source_File, values_to="p.value", names_to="pheno") %>%
  mutate(pheno = sub(".p$", "", pheno),
         sig_bfcor = case_when(pheno %in% vol_list_global & p.value < (0.05/length(vol_list_global)) ~ TRUE,
                            pheno %in% vol_list_global & p.value >= (0.05/length(vol_list_global)) ~ FALSE,
                            !(pheno %in% vol_list_global) & p.value < (0.05/length(ct_list)) ~ TRUE,
                            !(pheno %in% vol_list_global) & p.value >= (0.05/length(ct_list)) ~ FALSE,
                            TRUE ~ NA))
sex_ttest_long %>%
  group_by(Source_File) %>%
  summarize(n_sig_regions = sum(sig_bfcor),
            n_regions = n())
```

Basically same results gam/gamlss

#### Mu

Does the number of regions w significant age effects on mu change depending on combat version?

```{r}
drop1.df %>%
  dplyr::filter(Term=="poly(age_days, 3)" & Moment == "mu") %>%
  group_by(Dataset) %>%
  summarize(n_sig_regions = sum(sig.drop_bf.corr),
            n_regions = n())
```

No.

Does the number of regions w significant sex effects on mu change depending on combat version?

```{r}
drop1.df %>%
  dplyr::filter(Term=="sexMale" & Moment == "mu") %>%
  group_by(Dataset) %>%
  summarize(n_sig_regions = sum(sig.drop_bf.corr),
            n_regions = n())

sum.df %>%
  dplyr::filter(term=="sexMale" & parameter == "mu") %>%
  group_by(dataset) %>%
  summarize(n_sig_regions = sum(sig.sum_bf.corr),
            n_regions = n())
```

Slightly. Only combat w/o covariates or lm combat without a ref site are noticeably off. Consistent across drop1 and summary outputs

```{r}
sex.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & parameter == "mu" & pheno_cat != "Global Volume") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = t_stat)) +
  scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_void() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +
  ggtitle("t-statistic for effect of male sex on mu")
```

Trying with ultra-strict threshold to see if results change:

```{r}
drop1.df <- drop1.df %>%
  mutate(sig.sum.strict = case_when(drop1.pval < (0.05/length(pheno_list)) ~ TRUE,
                            drop1.pval >= (0.05/length(pheno_list)) ~ FALSE,
                            TRUE ~ NA))

drop1.df %>%
  dplyr::filter(Term=="poly(age_days, 3)" & Moment == "mu") %>%
  group_by(Dataset) %>%
  summarize(n_sig_regions = sum(sig.sum.strict),
            n_regions = n())
```

No difference

#### Sigma

```{r}
sex.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & parameter == "sigma" & pheno_cat != "Global Volume") %>%
  ggplot() +
  geom_col(aes(x=label, y=t_stat, fill=dataset)) +
  facet_wrap(~pheno_cat, scales = "free_x") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  #theme_void() +
  #theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +
  ggtitle("t-statistic for effect of male sex on sigma")

sex.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & parameter == "sigma" & pheno_cat != "Global Volume") %>%
  dplyr::filter(dataset == "cf.lm-data" | dataset == "cf.gamlss-data" | dataset == "ukb-CN-data-simsites") %>%
  ggplot() +
  geom_col(aes(x=label, y=t_stat, fill=dataset), position = "dodge") +
  facet_wrap(~pheno_cat, scales = "free_x") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  #theme_void() +
  #theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +
  ggtitle("t-statistic for effect of male sex on sigma")

sex.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & parameter == "sigma" & pheno_cat != "Global Volume") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = t_stat)) +
  scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_void() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +
  ggtitle("t-statistic for effect of male sex on sigma")
```
Check number of regions w significant sex effects in sigma
```{r}
sex.sigma.counts <- sex.df %>%
  dplyr::filter(term=="sexMale" & parameter == "sigma") %>%
  group_by(dataset, pheno_cat) %>%
  summarize(n_sig_regions = sum(sig.drop_bf.corr),
            n_regions = n())
sex.sigma.counts
```

More variation
```{r}
ggplot(sex.sigma.counts) +
  geom_col(aes(x=dataset, y=n_sig_regions, fill=dataset)) +
  facet_wrap(~pheno_cat) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

##### Variance
Convert sigma to variance by correcting for log-link function (sex's effect on variance = exp(sigma) - exp(log(sigma) - beta sex)) -> exp(beta_sex).
Also getting mean variance.

```{r}
sigma.df <- sum.df %>%
  dplyr::filter(parameter == "sigma") %>%
  mutate(var_est = exp(estimate),
         var_mean = exp(moment.mean),
         beta_ratio = var_est/var_mean)

skim(sigma.df$var_est)

sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & term =="sexMale") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_histogram(aes(x=var_est, fill = dataset), color="black", alpha=0.5, position = "identity") +
  theme_linedraw() +
  facet_wrap(~pheno_cat, scales= "free")
```

```{r, fig.height=10, fig.width=8}
proportion_beta_plot <- sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume" & term =="sexMale") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = beta_ratio)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=15) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Oslo") +
  ggtitle("Proportion of effect of male sex on variance")
```

```{r, fig.height=15, fig.width=12}
beta.plot <- sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume" & term =="sexMale") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = var_est)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=20) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Oslo") +
  ggtitle("Beta weight for effect of male sex on variance")

ggsave("male_variance_beta.jpeg", beta.plot, width=12, height=15)
```

```{r, fig.height=5, fig.widt=5}
sigma.df %>%
  dplyr::filter(pheno_cat == "Global Volume" & term =="sexMale") %>%
  mutate(var_est.sig = case_when(sig.sum_bf.corr == TRUE ~ var_est,
                                 sig.sum_bf.corr == FALSE ~ 0)) %>%
  #group_by(dataset) %>%
  ggplot() +
  geom_col(aes(x=dataset, y = var_est.sig, fill=dataset), position="dodge") +
  #scale_fill_gradient2() +
  facet_grid(~pheno) + 
  theme_bw(base_size=12) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position="none") +
  #theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  ggtitle("Beta weight for effect of male sex on global volumes' variance")

sigma.df.glob <- sigma.df %>%
  dplyr::filter(pheno_cat == "Global Volume" & term =="sexMale")
table(sigma.df.glob$dataset, sigma.df.glob$sig.sum_bf.corr)
```

fix age, get diff between male and female pred. variance
```{r}
pred.csvs <- list.files(path = "ukb_basic", pattern = "_variance.csv", full.names = TRUE)
combined_var <- data.frame()

for (file in pred.csvs) {
  # Read each CSV file
  data <- fread(file)

  # Add a "Source_File" column with the file name
  data <- data %>%
    mutate(Source_File = as.factor(basename(file))) %>%
    mutate(dataset = sub("_data_variance.csv", "", Source_File))

  # Bind the data to the combined dataframe
  combined_var <- bind_rows(combined_var, data, .id = "File_ID")
}

table(combined_var$dataset)
#skim(combined_preds)
```

```{r}
combined_var <- combined_var %>%
  mutate(sex_std = sex_effect/f.var)

summary(sigma.df$var_est)
summary(combined_var$sex_effect)

summary(sigma.df$beta_ratio)
summary(combined_var$sex_std)
```

```{r}
plot.info <- sigma.df %>%
  dplyr::filter(term == "sexMale") %>%
  dplyr::select(mod_name, pheno, dataset, pheno_cat, label, hemi, region, sig.sum_bf.corr) %>%
  unique() %>%
  mutate(dataset = gsub("-", "_", dataset),
         dataset = gsub("_simsites", "", dataset))
combined_var <- inner_join(combined_var, plot.info)
table(combined_var$dataset)
```

plot

```{r, fig.height=5, fig.widt=5}
combined_var %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = sex_std)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=12) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Oslo") +
  ggtitle("Standardized effect of male sex on variance")
```

using aseg

```{r}
plot(aseg)
```

## ComBat Effects on Centiles

Do centile scores change based on combat version?
```{r}
pred.csvs <- list.files(path = "ukb_basic", pattern = "_predictions.csv", full.names = TRUE)
combined_preds <- data.frame()

for (file in pred.csvs) {
  # Read each CSV file
  data <- fread(file)

  # Add a "Source_File" column with the file name
  data <- data %>%
    mutate(Source_File = as.factor(basename(file))) %>%
    mutate(dataset = sub("_data_predictions.csv", "", Source_File))

  # Bind the data to the combined dataframe
  combined_preds <- bind_rows(combined_preds, data, .id = "File_ID")
}

table(combined_preds$dataset)
#skim(combined_preds)
```
calc how much centiles change
```{r}
cent_diffs <- combined_preds %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ abs(. - first(.)), .names = "abs.diff_{.col}")) %>%
  ungroup()

diff_vol_glob <- sapply(vol_list_global, function(x) paste0("abs.diff_", x))

mean.cent.diffs <- cent_diffs %>%
  group_by(dataset) %>%
  summarize(across(starts_with("abs.diff_"), mean)) %>%
  ungroup()

max.cent.diffs <- cent_diffs %>%
  group_by(dataset) %>%
  summarize(across(starts_with("abs.diff_"), max)) %>%
  ungroup()
#names(max.cent.diffs)
```

Plot average differences in centiles
```{r}
mean.diffs.long <- mean.cent.diffs %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="centile_abs.diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)),
      label = sub("_[^_]*_", "_", pheno))
mean.diffs.long <- left_join(mean.diffs.long, dk.parc, by="label")
```

```{r}
mean.diffs.long %>%
  dplyr::filter(dataset != "ukb_CN") %>%
ggplot() +
  geom_density(aes(x=centile_abs.diff, fill=dataset), alpha=0.5) +
  facet_grid(dataset~pheno_cat)

mean.diffs.long %>%
  dplyr::filter(dataset== "cf.gam" | dataset == "cf.gam_refA" | dataset == "cf.gamlss" | dataset == "cf.gamlss_refA") %>%
  ggplot() +
  geom_density(aes(x=centile_abs.diff, fill=dataset), alpha=0.5) +
  scale_x_continuous(labels = label_comma())

mean.diffs.long %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(centile_abs.diff = centile_abs.diff*100) %>% #for visualization purposes
  ggplot() +
  geom_histogram(aes(x=centile_abs.diff, fill=dataset), color="black", binwidth=0.1, alpha=0.5, position="identity") +
  theme_linedraw() +
  #theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_x_continuous(labels = label_comma()) +
  ggtitle("Mean absolute differences in centile scores across ComBat \nprocessing pipelines for each phenotype")

cent.diffs.plot <- mean.diffs.long %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(centile_abs.diff = centile_abs.diff*100) %>% #for visuals
  ggplot() +
  geom_histogram(aes(x=centile_abs.diff, fill=dataset), color="black", alpha=0.6, position="identity") +
  theme_linedraw(base_size = 12) +
  scale_x_continuous(labels = label_comma()) + facet_wrap(~pheno_cat) +
  paletteer::scale_fill_paletteer_d("Polychrome::light") +
  ggtitle("Mean absolute differences in centile scores across \nComBat processing pipelines for each phenotype")
ggsave("mean_centile_diffs.jpeg", cent.diffs.plot)
```

```{r, fig.height=15, fig.width=12}
mean.diffs.long %>%
  dplyr::filter(pheno_cat != "Global Volume" & dataset != "ukb_CN") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = centile_abs.diff)) +
  scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_void() +
  theme_bw(base_size=20) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +
  ggtitle("Mean difference in centiles across ComBat configurations")
```

Look at max difference per phenotype
```{r, fig.height=6, fig.width=7}
max.diffs.long <- max.cent.diffs %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="max_centile_abs.diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

max.diffs.long %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(max_centile_abs.diff = max_centile_abs.diff*100) %>% #for visuals
  ggplot() +
  geom_histogram(aes(x=max_centile_abs.diff, fill=dataset), color="black", alpha=0.6, position="identity") +
  theme_linedraw(base_size = 12) +
  scale_x_continuous(labels = label_comma()) + facet_wrap(~pheno_cat) +
  paletteer::scale_fill_paletteer_d("Polychrome::light") +
  ggtitle("Maximum absolute differences in centile scores across \nComBat processing pipelines for each phenotype")
```

```{r}
max.diffs.long %>%
  mutate(max_centile_abs.diff = max_centile_abs.diff*100) %>% #for visuals
  dplyr::filter(dataset != "ukb_CN") %>%
  ggplot() +
  geom_histogram(aes(x=max_centile_abs.diff, fill=dataset), color="black", alpha=0.5, position="identity", binwidth = 0.5) +
  theme_linedraw(base_size = 12) +
  scale_x_continuous(labels = label_comma())+
  ggtitle("Maximum absolute differences in centile scores across \nComBat processing pipelines for each phenotype")
```

Test for differences in distribution
```{r}
#get centile errors for each subject
cent.diffs2 <- combined_preds %>%
  dplyr::filter(!str_detect(dataset, "_refA")) %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ abs(. - first(.)), .names = "diff_{.col}")) %>%
  ungroup() %>%
  dplyr::filter(dataset != "ukb_CN") #drop ref

# cf.gmvdiff <- cent.diffs2 %>%
#   dplyr::filter(dataset == "cf") %>%
#   dplyr::select(diff_GMV)
# cf.lm.gmvdiff <- cent.diffs2 %>%
#   dplyr::filter(dataset == "cf.lm") %>%
#   dplyr::select(diff_GMV) 
# 
# t.test(cf.gmvdiff, cf.lm.gmvdiff, paired=TRUE)
```

NOTE: "missing" paired comparisons between cf and cf.lm diffs are because diff values are identical between the two methods (i.e. there is literally no difference in accuracy of centile scores between the two)

```{r, warning=FALSE, message=FALSE}
#within each phenotype, iterate paired t test between each combo of combat dfs

#initalize empty dfs to store outputs
sum.df <- data.frame(pheno = character(),
                     term = character(),
                     df = double(),
                     sumsq = double(),
                     meansq = double(),
                     statistic = double(),
                     p.value = double())
t.df <- data.frame("pheno" = character(),
                   "group1" = character(),
                   "group2" = character(),
                   "p.value" = double())
pheno_diff.list <- paste0("diff_", pheno_list)

#run tests
attach(cent.diffs2)
for (pheno in pheno_diff.list) {
  df.aov <- tidy(aov(cent.diffs2[[pheno]] ~ cent.diffs2[["dataset"]]))
  df.aov$pheno <- pheno
  sum.df <- rbind(sum.df, df.aov)

  df.pairwise.t <- tidy(pairwise.t.test(x=cent.diffs2[[pheno]], g=cent.diffs2[["dataset"]], paired = TRUE, p.adj = "none"))
  df.pairwise.t$pheno <- pheno
  t.df <- rbind(t.df, df.pairwise.t)
}
detach()
```

```{r}
#apply fdr correction
sum.df <- sum.df %>%
  dplyr::filter(term != "Residuals") %>%
  mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(pheno_list)))

summary(sum.df$p.val_fdr)

t.df <- t.df %>%
  mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(pheno)),
         sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                             p.val_fdr >= 0.05 ~ FALSE),
         pheno = sub("diff_", "", pheno)) %>%
  unite(comp, c("group1", "group2")) %>%
  mutate(comp = as.factor(comp))
table(t.df$comp)

t.df %>%
  group_by(comp) %>%
   summarize(n_sig_regions = sum(sig_fdr),
            n_compare = n(),
            min_pval = min(p.val_fdr),
            max_pval = max(p.val_fdr))
```

#### Sex-Biases

Do the different combat methods introduce any kind of bias in centile scores?

```{r}
mean.cent.diffs.by.sex <- cent_diffs %>%
  group_by(dataset, sex) %>%
  summarize(across(starts_with("abs.diff_"), mean)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="centile_diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

mean.cent.diffs.by.sex %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(centile_diff = centile_diff *100) %>%
ggplot() +
  geom_density(aes(x=centile_diff, fill=sex), alpha=0.5) +
  facet_wrap(~dataset, scales = "free_y")

max.cent.diffs.by.sex <- cent_diffs %>%
  group_by(dataset, sex) %>%
  summarize(across(starts_with("abs.diff_"), max)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="max_centile_diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

max.cent.diffs.by.sex %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(max_centile_diff = max_centile_diff *100) %>%
ggplot() +
  geom_density(aes(x=max_centile_diff, fill=sex), alpha=0.5) +
  facet_wrap(~dataset, scales = "free_y")
```

No shift in abs. differences, looking at overall diff.
```{r}
#get centile errors for each subject - not abs val
cent.diffs3 <- combined_preds %>%
  dplyr::filter(!str_detect(dataset, "_refA")) %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ . - first(.), .names = "diff_{.col}")) %>%
  ungroup() %>%
  dplyr::filter(dataset != "ukb_CN") #drop ref
saveRDS(cent.diffs3, file="ukb_basic/cent_diffs.RDS")
```

```{r, eval=FALSE}
cent.diffs3 %>%
  #dplyr::filter(dataset == "cf.gamlss") %>%
  pivot_longer(starts_with("diff_"), names_to="pheno", values_to="diff", names_prefix = "diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_))) %>%
  ggplot() +
  geom_density(aes(x=diff, fill=dataset), alpha=0.4) +
  facet_wrap(~pheno_cat) +
  theme(legend.position="none")
```

```{r}
max.cent.diffs.by.sex2 <- cent.diffs3 %>%
  group_by(dataset, sex) %>%
  summarize_at(vars(starts_with("diff")), list(max_value = ~.[which.max(abs(.))])) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("diff_"), values_to="max_centile_diff", names_to = "pheno", names_prefix = "diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

max.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(max_centile_diff = max_centile_diff *100) %>%
ggplot() +
  geom_density(aes(x=max_centile_diff, fill=sex), alpha=0.5) +
  facet_wrap(~dataset, scales="free")
```

Other methods are fairly consistently underestimating male centiles and overestimating female centiles.

```{r}
mean.cent.diffs.by.sex2 <- cent.diffs3 %>%
  group_by(dataset, sex) %>%
  summarize(across(starts_with("diff_"), mean)) %>%
  ungroup() %>%
  pivot_longer(cols=starts_with("diff_"), values_to="centile_diff", names_to = "pheno", names_prefix = "diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

mean.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  mutate(centile_diff = centile_diff *100) %>%
ggplot() +
  geom_density(aes(x=centile_diff, fill=sex), alpha=0.5) +
  facet_wrap(~dataset, scales="free")
```

T test of mean diff. between sex w/in each dataset
```{r}
t.test.by.sex <- mean.cent.diffs.by.sex2 %>%
  dplyr::filter(dataset != "ukb_CN") %>%
  group_by(dataset) %>%
  summarise(t_statistic = t.test(centile_diff ~ sex)$statistic,
            p_value = t.test(centile_diff ~ sex)$p.value) %>%
  mutate(p_val_fdr.corr = p.adjust(p_value, method="fdr", n = length(dataset)),
         fdr_sig = case_when(p_val_fdr.corr < 0.05 ~ TRUE,
                             p_val_fdr.corr >= 0.05 ~ FALSE))

t.test.by.sex
```

Both gam methods and lm combat with ref both end up inducing a bias in how males and females centiles are shifted.

```{r, warning=FALSE, message=FALSE}
# within each df, within each phenotype, t-test between male and female diffs

#initalize empty dfs to store outputs
t.df.sex <- data.frame("pheno" = character(),
                   "dataset" = character(),
                   "p_value" = double()
                   )

>>>>

---
title: "UKB Raw Analyses"
output: html_document
date: '2023-10-06'
---

Mapping sex effects in raw, un-combatted data in order to establish deviation induced by ComBat-ing on simulated sites.

@@

base_path <- "/Users/megardn/Desktop/BGD_Repos/combat_biovar/"

source(paste0(base_path, "cubic_scripts/R_scripts/gamlss_helper_funs.R"))
# 
# # Enable this universe
# options(repos = c(
#   ggseg = 'https://ggseg.r-universe.dev',
#   CRAN = 'https://cloud.r-project.org'))
# 
# # Install some packages
# install.packages('ggsegExtra')
```

@@

```{r}
ukb_df_totals <- read_csv(file=paste0(base_path, "data/ukb_CN_data.csv"))
ukb_df_age.filt <- read_csv(file=paste0(base_path, "data/ukb_to_model/ukb_CN_data_simsites.csv"))

vol_list_global <- readRDS(file=paste0(base_path, "cubic_scripts/R_scripts/vol_list_global.rds"))
vol_list_regions <- readRDS(file=paste0(base_path, "cubic_scripts/R_scripts/Vol_list_regions.rds"))
sa_list <- readRDS(file=paste0(base_path, "cubic_scripts/R_scripts/SA_list.rds"))
ct_list <- readRDS(file=paste0(base_path, "cubic_scripts/R_scripts/CT_list.rds"))
pheno_list <- readRDS(file=paste0(base_path, "cubic_scripts/R_scripts/pheno_list.rds"))

#aseg_cerebrum <- readRDS(file="/Users/megardn/Desktop/BGD_Repos/ggseg_atlas/aseg_cerebrum_atlas.RDS")
```

@@

```{r}
drop1.df <- fread(paste0(base_path, "ukb_basic/drop1_tests.csv"), stringsAsFactors = TRUE) %>%
  dplyr::select(-V1)
sex.df <- fread(paste0(base_path, "ukb_basic/sex_summary.csv"), stringsAsFactors = TRUE) %>%
  dplyr::select(-V1)
sum.df <- fread(paste0(base_path, "ukb_basic/gamlss_summary.csv"), stringsAsFactors = TRUE) %>%
  dplyr::select(-V1) %>%
  mutate(dataset = gsub("-data", "", dataset)) %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb-CN-simsites"))
```

@@

## Raw Data Results

@@
